The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2019-07-01 razzo-abstract.tex
Updates from Overleaf

% One or two sentences providing a basic
% introduction to the field,
% comprehensible to a scientist in any discipline.
There exist millions of species on Earth,
all originating from a common ancestor billions
of years ago.
The field of phylogenetics uses heritable material
to determine which species are closest related and what are
the mathematics that shape speciation.

% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
In Bayesian phylogenetics, a DNA/RNA/protein alignment
is used to infer a distribution of phylogenies and parameter estimates.
To do so, we use assumptions that may be biologically unrealistic,
but may give tolerable errors.

% One sentence clearly stating the general
% problem being addressed by this particular
% study.
Contemporary inference assumes that speciation
never co-occurs.

% One sentence summarising the main
% result (with the words “here we show”
% their equivalent).
Here we show the error we make in our inference,
when nature has varying degrees of co-occurring speciation.

% Two or three sentences explaining what
% the main result reveals in direct
% comparison to what was thought to be the case
% previously, or how the main result adds to
% previous knowledge.

% One or two sentences to put the results into a
% more general context.

% Two or three sentences to provide a
% broader perspective, readily comprehensible
% to a scientist in any discipline, may be included
% in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced
% by their inclusion. Under these circumstances, the length of the paragraph can be up to 300 words.
\end{abstract}
{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior

>>>>

% One or two sentences providing a basic
% introduction to the field,
% comprehensible to a scientist in any discipline.
There exist millions of species on Earth,
all originating from a common ancestor billions
of years ago.
The field of phylogenetics uses heritable material (e.g. DNA)
to determine the evolutionary history of species.

% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
Starting from heritable material and explicit assumptions,
Bayesian phylogenetics allows to infer a jointly-estimated 
phylogeny and parameter estimates distribution.
One of these assumptions is the speciation model, which
mathematically describes the branching process of a phylogeny in time.
Speciation models commonly assume that speciation events are independent and disjointed. Yet, such assumption may overlook the complexity of certain processes, where environmental changes promote speciation events in multiple lineages.
% Purpose of the paper
This new layer of complexity can be captured developing a novel ad hoc speciation model. However, the introduction of a new model could be not necessary if current models are capable of describing the process, under simpler assumptions.
Here we investigate the extent of the discrepancy produced by current BEAST2's tree priors, when trying to infer back trees where speciation can co-occur.

\iffalse
This research answers the question what the impact is of using a
species tree model that assumes speciation is independent, when it
is used on phylogenies created by a tree model in which speciation can
co-occur.
\fi

% Summarising the main result
We let the impact of co-occurring speciation vary on datasets of simulated trees and show the corresponding error produced during the inference process.

% Two or three sentences explaining what
% the main result reveals in direct
% comparison to what was thought to be the case
% previously, or how the main result adds to
% previous knowledge.
We show that the extent of the inference error grows with the amount of co-occuring speciation events to establish the limits of BEAST2's standard tree priors.

% One or two sentences to put the results into a
% more general context.
These results allow phylogeneticists to judge under which circumstances the commonly used speciation model can be safely used.

\end{abstract}

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior,
pirouette, BEAST2, babette

### 2019-07-01 razzo-methods.tex
Put 'tab_simparameters.tex' in root folder for thesis
Updates from Overleaf

@@

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates present 
also in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
$\nu$ is the rate at which an environmental change is triggered.
When such event is triggered, each species present in the phylogeny at that 
moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature respect to the one triggered 
by $\lambda$. In fact, whereas parameter $\lambda$ can be seen as describing 
a sympatric process, $\nu$ induces the rise of geographical barriers 
interrupting the gene flow and leading to an allopatric speciation.
Even though multiple speciations can co-occur, polytomies are not allowed 
in such process as each species can speciate only once at the time.
A likelihood expression for the process is provided in \cite{mbd}.

@@

We simulate the speciation process in continuous time using the 
Doob-Gillespie algorithm, using 
the \verb;mbd_sim; function in the \verb;mbd; package (\citep{mbd}).
We let parameters vary using all possible combinations of values
as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, we generate $1000$ independent 
reconstructed trees of the same crown age.
We have picked the parameters in such a way that in the most speciose
setting, the simulated trees have usually less than 200 taxa.

@@

From each MBD tree, we measure the impact 
of using the simpler BD tree prior, 
using the \verb;pirouette; R package, as described
in detail in \citep{pirouette}.

In brief: \verb;pirouette; starts from a 'true' (but 
unobservable in nature) starting phylogeny,
from which a DNA sequence alignment (which is observable 
in nature) is simulated. 
From each sequence alignment, a Bayesian inference is run, 
to obtain a posterior distribution of jointly-estimated 
trees and model parameter estimates.
By comparing the true tree and the posterior trees,
an inference error distribution is generated. 
We use the twinning option available in \verb;pirouette;
to measure a minimum and full error, in which
the minimum error is caused
by stochasticity in the full pipeline, where the full error
is the added error from using an invalid but standard speciation model.

In our context, the alignments are 1000 nucleotides in length,
with a known root sequence of four 250 mono-nucloetide blocks,
following the simplest nucleotide substitution model (JC69)
and clock model (strict), with a mutation rate of $\frac{1}{2}\cdot c$, 
in which $c$ is the crown age. With this mutation rate, each nuclotide is
expected to mutate (both silently and non-silently) 
in half of the histories from the root sequence in the past 
to the sequence in the tips in the present.

@@

For the error measurement, we use the nLTT statistic and
a burn-in of 0.1.

>>>>

We proceed in the following way: we build simulated datasets generated under the multiple birth model. Then we run a \verb;pirouette; analysis, which will lead to error distributions between the inference posterior and the original simulated trees. Importantly, the \verb;pirouette; analysis includes also a 'twin' parallel pipeline, which will provide a measure of the baseline error due to pure stochasticity, unavoidably occurring in the process.

@@

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, to the common per-species speciation and extinction rates already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.
When such event is triggered, each species present in the phylogeny at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature respect to the one triggered by $\lambda$. In fact, whereas parameter $\lambda$ can be seen as describing a sympatric process, $\nu$ induces the rise of geographical barriers interrupting the gene flow \giovanni{@richel: please check if biology here is accurate} and leading to an allopatric speciation.
Even though multiple speciations can co-occur, polytomies are not allowed in such process as each species can speciate only once at the time.
A likelihood expression for the process is provided in \cite{mbd}.

@@

We simulate the speciation process in continuous time using the Doob-Gillespie algorithm, using the \verb;mbd_sim; function from the \verb;mbd; package (\citep{mbd}).
We let parameters vary using all possible combinations of values as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, we generate $1000$ independent reconstructed trees of the same crown age.
\giovanni{Do we need to add more information on the Doob-Gillespie algorithm or it is overkilling?}

@@

From each MBD tree, we measure the impact of using the simpler BD tree prior on the inference, using the \verb;pirouette; R package, as described in detail in \citep{pirouette}.

In brief: \verb;pirouette; starts from a 'true' (but unobservable in nature) starting phylogeny, from which a DNA sequence alignment (which is observable in nature) is simulated. 
From each sequence alignment, a Bayesian inference is run, to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, an inference error distribution is generated. 
We use the twinning option available in \verb;pirouette; to measure a minimum and full error, in which the minimum error is caused by stochasticity in the full pipeline, where the full error is the added error from using an invalid but standard speciation model \giovanni{I am not sure I get this. What do you mean?}.

In our context, the alignments are 1000 nucleotides in length, with a known root sequence of four 250 mono-nucleotide blocks, following the simplest nucleotide substitution model (JC69) and clock model (strict), with a mutation rate of $\frac{1}{2}\cdot c$,  in which $c$ is the crown age. With this mutation rate, each nucleotide is expected to mutate (both silently and non-silently) 
in half of the histories from the root sequence in the past to the sequence in the tips in the present.

@@

For the error measurement, we use the nLTT statistic \citet{janzen2015} and
a burn-in fraction of $10\%$.

### 2019-07-01 razzo-results.tex
Updates from Overleaf

The inference error made for each of the parameter combinations
is shown in Fig.~\ref{fig:results}. For both extinction rates,
we find, as expected, that the error increases for increased $\nu$
or $q$. Also in line with our predictions, we find no difference
between the two extinction rates.

>>>>

The inference error made for each of the parameter combinations
is shown in Fig.~\ref{fig:results}. For both extinction rates,
we find that the error increases as $\nu$
and $q$ increase. Also in line with our predictions, we find no difference
between the two extinction rates \giovanni{I don't actually have predictions on that.}.
